Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics\\Black swan ecological dynamics\\ \ldots

### 2014-10-17 ms.tex
Spell check and clean up TeX

Our main analysis focusses on the commonly applied phenomenological Gompertz population dynamics model. In previous analyses of the GPDD, the Gompertz has proved to be the most frequently parsimonious model \citep{brook2006}. We extended the traditional Gompertz model by allowing the process noise to be drawn from a Student-t distribution with degrees of freedom parameter $\nu$ (Fig.~\ref{fig:didactic}). If $\nu$ is small ($\lesssim 10$) the t distribution has much heavier tails than a normal distribution. For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $1.8$\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution. By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are (Fig.~\ref{fig:didactic}).

@@

Our results clarify, but do not disagree with previous related analyses. Previous work with an older version of the GPDD assessed heavy-tailed behaviour but focused on identifying if the lognormal distribution was the most frequently parsimonious model of population abundance \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. Furthermore, by probabilistically quantifying process noise tail behaviour on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could create falsely generate heavy tails. This rebutal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with inverse-gamma-distributed variances,][]{gelman2014}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
(a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
Small values of $\nu$ create heavy tails.
As $\nu$ approaches infinity the distribution approaches the normal distribution.
For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
(c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
Coloured dots in panels c and d represent jumpts with less than a 1 in 1000 chance of occurring in a normal distribution.
(f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
Colour shading behind panels f--h illustrates the region of heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

>>>>

Our main analysis focuses on the commonly applied phenomenological Gompertz population dynamics model. In previous analyses of the GPDD, the Gompertz has proved to be the most frequently parsimonious model \citep{brook2006}. We extended the traditional Gompertz model by allowing the process noise to be drawn from a Student-t distribution with degrees of freedom parameter $\nu$ (Fig.~\ref{fig:didactic}). If $\nu$ is small ($\lesssim 10$) the t distribution has much heavier tails than a normal distribution. For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $1.8$\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution. By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are (Fig.~\ref{fig:didactic}).

@@

Our results clarify, but do not disagree with previous related analyses. Previous work with an older version of the GPDD assessed heavy-tailed behaviour but focused on identifying if the lognormal distribution was the most frequently parsimonious model of population abundance \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. Furthermore, by probabilistically quantifying process noise tail behaviour on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could create falsely generate heavy tails. This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with inverse-gamma-distributed variances,][]{gelman2014}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
(a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
Small values of $\nu$ create heavy tails.
As $\nu$ approaches infinity the distribution approaches the normal distribution.
For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
(c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
Coloured dots in panels c and d represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution.
(f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
Colour shading behind panels f--h illustrates the region of heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

### 2014-10-17 supp.tex
Spell check and clean up TeX

\begin{centering}
\LARGE
Supporting Material for TODO INSERT TITLE HERE\\[1.5em]
\end{centering}

@@

We followed the following data selection and quality-control rules with the GPDD:

@@

Our prior on $\nu$ is based on \citet{fernandez1998}; they chose a more informative exponential rate parameter of 0.1. We chose a less informative rate parameter of 0.01 and truncated the distribution at two, since at $\nu < 2$ the variance of the t distribution is undefined. This prior gives only a 7.7\% probability $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be sufficiently informative about the precise value of $\nu$. In the scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h) the posterior will approximately match the prior (median $= 71$, mean $= 102$) and not be flagged as likely heavy tailed using the metrics we used in our paper (e.g.~Pr($\nu < 10$) \textgreater{} 0.5).

We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}. We began with four chains and 2000 iterations, discarding the first 1000 as warm up (i.e.~4000 total samples). If $\hat{R}$ (a measure of chain convergence) was greater than 1.05 for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than 200, we doubled both the total iterations and warm up period and sampled from the model again. These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals. In the majority of cases our minimum thresholds were greatly exceeded. We continued this procedure up to 8000 iterations (16000 total samples) by which all chains were deemed to have sufficiently converged.

@@

First, we tested the ability to estimate $\nu$ given different true values of $\nu$ and sample sizes. We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$. We started with 1600 stochastic draws and then fitted the models again at the first 800, 400, 200, 100, 50, and 25 draws. Each time we recorded the posterior samples of $\nu$.

@@

%2 parts: how many samples from the true population t distribution do you need to detect low nu? And, given that you have a set of deviations in which nu is detectable (effective nu is within 0.5 CV of true nu), can the more complex Gompertz still capture this?

In the second part of our simulation testing, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates given that process deviations were provided in which our simpler model (in the previous section) could accurately estimate the value of $\nu$. To generate these process errors, we repeatedly generated candidate process deviations and estimated the central tendency, scale, and $\nu$ values each time. We recorded when the estimated nu was within $0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this set of random seed values in our Gompertz simulation. We then fitted our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD. We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for (correctly) specified observation error.

@@

%Fig.~\ref{fig:sim-gompertz}
%Fig.~\ref{fig:sim-gompertz-boxplots}
%Fig.~\ref{fig:sim-prob}

%Simulation testing: the ability to recover $\nu$ when randomly sampling from various distributions (Fig.~\ref{fig:sim-nu}); heavy-tailed Gompertz model performance and confidence interval coverage given the process noise has deviations that are known to have effective $\nu$ equal to true $\nu$ (Fig.~\ref{fig:sim-gompertz}); boxplots of the same output (Fig.~\ref{fig:sim-gompertz-boxplots}); probability that $\nu < 10$ for the Gompertz simulation testing (Fig.~\ref{fig:sim-prob}).

>>>>

\begin{centering}
\LARGE
Supporting Material\\[1.0em]
\end{centering}

@@

We applied the following data selection and quality-control rules to the Global Population Dynamics Database (GPDD):

@@

Our prior on $\nu$ is based on \citet{fernandez1998}; they chose a more informative exponential rate parameter of $0.1$. We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined. This prior gives only a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be sufficiently informative about the precise value of $\nu$. In the scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h) the posterior will approximately match the prior (median $= 71$, mean $= 102$) and not be flagged as likely heavy tailed using the metrics we used in our paper (e.g.~Pr$(\nu < 10) > 0.5$).

We fitted our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}. We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again. These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals. In the majority of cases our minimum thresholds were greatly exceeded. We continued this procedure up to $8000$ iterations ($16000$ total samples) by which all chains were deemed to have sufficiently converged.

@@

First, we tested the ability to estimate $\nu$ given different true values of $\nu$ and sample sizes. We took stochastic draws from t distributions with different $\nu$ values ($\nu = 3, 5, 10,$ and $10^6$ [$\approx$ normal]), with central tendency parameters of $0$, and scale parameters of $1$. We started with $1600$ stochastic draws and then fitted the models again at the first $800, 400, 200, 100, 50,$ and $25$ draws. Each time we recorded the posterior samples of $\nu$.

@@

In the second part of our simulation testing, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates given that process deviations were provided in which our simpler model (in the previous section) could accurately estimate the value of $\nu$. To generate these process errors, we repeatedly generated candidate process deviations and estimated the central tendency, scale, and $\nu$ values each time. We recorded when $\hat{\nu}$ was within $0.2$ CVs (coefficient of variations) of the true $\nu$ value and used this set of random seed values in our Gompertz simulation. We then fitted our Gompertz models to the simulated datasets with all parameters (except $\nu$) set near the median values estimated in the GPDD. We repeated this with $50$ and $100$ samples without observation error, $50$ samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples with the same observation error and a Gompertz model that allowed for correctly specified observation error magnitude.

@@
